---
title: "Tidy Data Demo"
author: "Stephen Piccolo"
output: html_document
---

### Load libraries & set working directory

```{r}
library(reshape2)
library(stringr)
library(readr)
library(dplyr)

setwd("~/Dropbox/Teaching/BIO465")
```

### Load Pew religion data

This data set shows the relationship between religion and income (surveyed by the Pew Research Center).

```{r}
rawPewData <- read_tsv("TidyData/pew2.txt")
head(rawPewData, n=20)
```

### "Melt" the income columns into two columns

```{r}
tidyPewData <- melt(rawPewData, "religion") # reshape2 package
head(tidyPewData, n=20)
```

### Load tuberculosis data

This data set represents the number of cases of tuberculosis observed in a WHO study.

```{r}
rawTbData <- read_csv("TidyData/tb2.csv")
head(rawTbData, n=20)
```

### "Melt" the sex/age variable and remove cases with missing values

```{r}
tb <- melt(rawTbData, id = c("country", "year"), na.rm = TRUE)
head(tb, n=20)
```

### Split apart the sex and age values into their own columns

```{r}
# Extract sex from composite variable
tb$sex <- str_sub(tb$variable, 1, 1)

# Extract age from composite variable
tb$age <- str_sub(tb$variable, 2)

# Remove composite "variable" column
tb$variable <- NULL

head(tb, n=20)
```

### Reformat age values for readability

```{r}
ageOptions <- c("1524" = "15-24", "2534" = "25-34", "3544" = "35-44", "4554" = "45-54", "5564" = "55-64", "65" = "65+")
tb$age <- ageOptions[tb$age]
head(tb, n=20)
```

### Select columns

##### (In this case, we are selecting all the columns but reordering them)

```{r}
selected <- select(tb, country, year, sex, age, value)
head(selected, n=20)
```

### Sort by specific column values

```{r}
reordered <- arrange(selected, country, year, sex, age)
head(reordered, n=20)
```

### Filter

```{r}
filtered <- filter(selected, country=="AD")
head(filtered, n=20)
```

### Identify distinct values for a given column

```{r}
distinct(select(tb, country))
```

### Summarize a single column

```{r}
summarise(tb, average=mean(value))
```

### Count total per country

```{r}
gb <- group_by(tb, country)
summarise(gb, count=n())
```

### Count total per country per age group

```{r}
gb <- group_by(tb, country, age)
summarise(gb, count=n())
```

### Max per country per age group

```{r}
gb <- group_by(tb, country, age)
summarise(gb, count=n(), max=max(value))
```

### Add a new column

```{r}
tb2 <- mutate(tb, valueSquared=value^2)
head(tb2, n=20)
```

### Inner Join (simple)

```{r}
data1 <- data.frame(samples=paste("Sample", 1:10, sep=""), Gene1=rnorm(10), Gene2=rnorm(10))
data2 <- data.frame(samples=paste("Sample", 1:10, sep=""), Gene3=rnorm(10), Gene4=rnorm(10))
data3 <- inner_join(data1, data2, by="samples")

print(data1)
print(data2)
print(data3)
```

### Inner Join (with non-overlapping samples)

```{r}
data1 <- data.frame(samples=paste("Sample", 1:10, sep=""), Gene1=rnorm(10), Gene2=rnorm(10))
data2 <- data.frame(samples=paste("Sample", 6:15, sep=""), Gene3=rnorm(10), Gene4=rnorm(10))
data3 <- inner_join(data1, data2, by="samples")

print(data1)
print(data2)
print(data3)
```

### Left Join (with non-overlapping samples)

```{r}
data1 <- data.frame(samples=paste("Sample", 1:10, sep=""), Gene1=rnorm(10), Gene2=rnorm(10))
data2 <- data.frame(samples=paste("Sample", 6:15, sep=""), Gene3=rnorm(10), Gene4=rnorm(10))
data3 <- left_join(data1, data2, by="samples")

print(data1)
print(data2)
print(data3)
```

### Full Join (with non-overlapping samples)

```{r}
data1 <- data.frame(samples=paste("Sample", 1:10, sep=""), Gene1=rnorm(10), Gene2=rnorm(10))
data2 <- data.frame(samples=paste("Sample", 6:15, sep=""), Gene3=rnorm(10), Gene4=rnorm(10))
data3 <- full_join(data1, data2, by="samples")

print(data1)
print(data2)
print(data3)
```